# Appendices {#app:appendix}
`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'`


```{r testtab}
d <- structure(list(
  Year = c("2018", "2019", "2020"), 
  `Value 1` = c(1.12, 2.32, 3.67), 
  `Value 2` = c(31.9, 2.8, 112.2)), 
  row.names = c(NA, -3L), class = "data.frame")

csasdown::csas_table(d,
  align = c("c", "r", "r"),
  caption = "Test table with a caption will be numbered automatically.")
```


## Example Objectives/Performance Metrics from Other Jurisdictions

Example 1 – Viability and Risk Assessment Procedure (VRAP) for Puget Sound Chinook salmon (NOAA 2017)

VRAP is a process implemented by the National Marine Fisheries Service, NMFS, to calculate exploitation rates for Pacific Salmon in the US based on upper and lower benchmarks (Rebuilding Escapement Threshold (RET) and Critical Escapement Threshold (CET), respectively). For each population, a RET is defined as spawner abundance at MSY (if stock-recruitment data are available), or an absolute abundance derived from the conservation literature in the absence of data, e.g., 1250 spawners for Puget Sound Chinook Salmon (NOAA 2017). A CET is defined as 5% of equilibrium spawner abundance, or an absolute abundance derived from the conservation literature, e.g., 200 spawners for Puget Sound Chinook Salmon. The rebuilding exploitation rate is set by evaluating performance over a range of exploitation rates and selecting the largest exploitation rate where less than 5% of simulation runs are below CET over 25 years, and the smallest exploitation rate where greater than 80% of runs are above RET over 25 years.

Example 2 – US rebuilding plans for federally regulated fisheries under the Magnuson-Stevens Act (National Research Council 2014)

For federally regulated fisheries in the US (including Pacific Salmon) the Magnuson-Stevens Acts legally mandates the implementation of rebuilding plans for overfished stocks. NFMS guidelines for rebuilding describe the requirements based on: 

- A target stock size

- A minimum stock size, below which rebuilding plans are required (e.g., for Pacific salmon, spawning biomass < 0.5 or 0.75 MSY levels; National Research Council 2014)

- A minimum acceptable  probability of rebuilding to the target, set by management

- Target time for recovery (bounded by minimum and maximum times for recovery), where these times are specified within a closed-loop stock dynamics model in the absence of fishing

Rebuilding analyses are comprised of projections in a closed-loop simulation model for a range of fishing mortality rates to identify those that achieve rebuilding targets with an acceptable probability within the maximum time frame. A set of decision rules are implemented if stocks fail to rebuild in order to revise targets and fishing mortality rates within the rebuilding plan (Wetzel and Punt 2016).

Example 3- The Committee on the Status of Endangered Wildlife in Canada, COSEWIC 

COSEWIC has identified critical thresholds on abundances to delineate conservation risks for Endangered and Threatened populations at 250 and 1000 mature individuals, or 2500 and 10,000 mature individuals when combined with declining trajectories, respectively. COSEWIC further identified thresholds on probabilities of extirpation for Endangered and Threatened categories, at 10% over 100 years and 20% over 5 generations or 20 years, whichever is longer.  These thresholds are analogous to those used by IUCN and are derived from conservation literature on population dynamics at small population sizes across terrestrial and aquatic species.



```{r CandThresh}
dat <- tribble(
  ~`Category`, ~`Description`, ~`Threshold (Reference)`,
  "Wild Salmon Policy lower benchmark*", "Level which provides a substantial buffer between it and levels that would lead to CU being considered at risk of extirpation by COSEWIC",
  "\\emph{Sgen}, spawner abundances that will recover to \\emph{SMSY} within 1 generation in the absence of fishing under equilibrium conditions  (1)", 
  " ", " ",
  "25th percentile of observed spawner abundances (2)",
  " ", " ",
  "20\\% of spawners at maximum recruitment, \\emph{Smax} (1,3,4)",
  "COSEWIC thresholds*", "Endangered (Criterion D)", "250 (5)",
  " ", "Threatened (Criterion D)", "1000 (5)",
  " ", "Endangered, when combined with continuing decline (Criterion C)", "2500 (5)",
  " ", "Threatened, when combined with continuing decline (Criterion C)", "10,000 (5)",
  "Pacific Fisheries Management Council, US", 
  "Definition of “overfished” for Pacific salmon, below which rebuilding plans must be developed",
  "0.5\\emph{BMSY}  or 0.75\\emph{BMSY} (6)",
  "NMFS, Viability and Risk Assessment Procedure (VRAP) for computing Rebuilding Exploitation Rates (RERs)",
  "Critical escapement threshold (CET), represents a boundary below which uncertainties about population dynamics increase substantially related to genetic and environmental risks",
  "5\\% equilibrium spawner abundances, or 42-417 annual spawners (7)",
  " ", " ", "200 annual spawners (for Puget Sound Chinook Salmon) (7,8)",
  "Cultus Lake Sockeye Salmon Recovery Objectives", "Ensure the genetic integrity of the population ",
  "1000 successful adult spawners, averaged over a generation, with no fewer than 500 in any one year (9)",
  "Sakinaw Lake Sockeye Recovery Objective", "Safeguard genetic diversity with interim milestones",
  "500 natural adult spawners after 2 generations and 1000 after 4 generations (10)",
  "Interior Fraser Coho Salmon Recovery Objectives", 
  "Secure the long term viability and diversity of naturally spawners", 
  "1000 natural spawners in at least half of sub-populations, averaged over 3 years (11)"
  
  )
  

dat %>% 
  mutate_all(kableExtra::linebreak) %>%
  csas_table(format = "latex", escape = FALSE, font_size = 11, linesep = "\\addlinespace",
             caption = "Candidate lower rebuilding thresholds and examples from case studies") %>%
  #kableExtra::row_spec(4, hline_after = TRUE) %>% 
  kableExtra::column_spec(1, width = "10em") %>%
  kableExtra::column_spec(2, width = "15em") %>%
  kableExtra::column_spec(3, width = "15em")
 
             
             

```


\footnotesize *WSP and COSEWIC assess number of mature individuals as the geometric mean spawner abundance over the most recent generation to compare against thresholds. In contrast, other assessments have compared similar thresholds against total population size calculated as the product of the mean spawner abundances × the average generation length [@allendorf1997] to assess risks of extinction related to genetic effects. NOAA [-@noaa2017] reported in fish/year.
(1) [@holt2009]  
(2) [@holt2018]  
(3) [@johnston2002]  
(4) [@shortreed2001]  
(5) [@cosewic2015]  
(6) [@nrc2014]  
(7) [@mcelhaney2000]  
(8) [@noaa2017]  
(9) [@cultus2009]  
(10) [@sakinaw2005]




```{r CandExt}
dat <- tribble(
  ~`Category`, ~`Description`, ~`Threshold (Reference)`,
    "From the conservation literature", "Quasi-extirpation threshold for salmon", 
  "100 fish (50-250) for four consecutive years (1,2)",
  "NMFS, Viability and Risk Assessment Procedure (VRAP) for computing Rebuilding Exploitation Rates (RERs)",
  "Quasi-extirpation threshold for salmon", "250 fish/generation (3)",
  "COSEWIC definitions for categories for extirpation", "Threatened",
  "10\\% chance of extirpation over 100 years (3)", 
  " ", "Endangered", "20\\% chance of extirpation within the longer of 5 generations or 20 years"
  )
  

dat %>% 
  mutate_all(kableExtra::linebreak) %>%
  csas_table(format = "latex", escape = FALSE, font_size = 11, linesep = "\\addlinespace",
             caption = "Candidate extirpation thresholds and examples from case studies") %>%
  #kableExtra::row_spec(4, hline_after = TRUE) %>% 
  kableExtra::column_spec(1, width = "10em") %>%
  kableExtra::column_spec(2, width = "15em") %>%
  kableExtra::column_spec(3, width = "15em")
 
             
             

```

```{r footnotetest}
x <- knitr::kable(head(mtcars), "latex")
add_footnote(x, c("footnote 1", "footnote 2"), notation = "number")
```

```{r ExamPMs}
dat <- tribble(
  ~`Overarching goal`, ~`Performance metrics`,
  "Rebuild depleted CUs\n(These performance metrics infer a relatively monotonic trajectory towards a rebuilding target)",
   "Probability that any one (or a specified proportion of) red-status CU(s) rebuilds to above its lower rebuilding threshold within a given time frame.", 
  "", "Probability that all red-status CUs rebuild to above their lower rebuilding thresholds (e.g., lower WSP benchmark) within a given time frame.",
  " ","Proportion of red-status CUs that rebuild to above their lower rebuilding thresholds with a specified probability within a given time frame. (Median over MC trials and 95\\% CIs)", 
  " ", "Number of years required to achieve lower rebuilding thresholds for one (a proportion of, or all) red-listed CU(s) with a specific probability. (Median over MC trials and 95\\% CIs)", "","",
  
  "Minimize risk of loss\n(These metrics can be duplicated for numerous rebuilding and target thresholds)", 
  "Proportion of years that all CUs are above a quasi-extirpation threshold across the modelled time period. (Median over MC trials and 95\\% CIs)", 
  "", "Proportion of years where all CUs are above their lower rebuilding thresholds within the modelled time period. (Median over MC trials and 95\\% CIs)",
  "", "Proportion of years where at least one (or a specified \\% of) CU(s) is(are) above their lower rebuilding thresholds within the modelled time period. (Median over MC trials and 95\\% CIs)",
  "", "Proportion of years where all CUs remain above their lower (or upper) rebuilding threshold across the modelled time period. (Median over MC trials and 95\\% CIs)",
  "", "Mean spawner abundances over the modelled time-period (or most recent generation) relative to lower rebuilding threshold. (Median over MC trials and 95\\% CIs)",
  "", "Variation in spawner abundances: CV of (or average \\% change between years in) spawner abundances over the modelled time period. (Median value over MC trials and 95\\% CIs) (suggested as indicator of extinction risk by Wainwright and Waples 1998)"," "," ",
  
  "Avoid COSEWIC listing", "Short-term trends in spawner biomass over the last three generations. (Median over MC trials and 95\\% CIs) \\emph{COSEWIC Criterion A}",
  "", "Probability that short-term trends in abundances are > 30\\% (COSEWIC threshold) in the most recent time period. \\emph{COSEWIC Criterion A}",
  "", "Proportion of years where the short-term trend metrics < 30\\% for all CUs (or specified \\% of CUs). (Median over MC trials and 95\\% CIs) \\emph{COSEWIC Criterion A}",
  "", "Proportion of years where the short-term trend is stationary or positive and abundances are greater than 10,000. \\emph{COSEWIC Criterion C}",
 "", "Proportion of years where all CUs are above COSEWIC small population size thresholds, across the entire sampling period. (Median over MC trials and 95\\% CIs). \\emph{COSEWIC Criterion D}"," "," ",

  "Maintain exploitation rates below sustainable levels", 
  "Mean exploitation rate relative to current $U_{MSY}$ for over the modelled time period. (Median value over MC trials and 95\\% CIs)"," "," ",

  "Maximize catch and stability in catch", 
  "Proportion of years that mean catch for the CU-aggregate is above a minimum acceptable level over the entire sampling period, in the short term (first 1-2 generations), or in the long term (last 1-2 generations). (Median value over MC trials and 95\\% CIs)", 
  "", "Mean catch over the entire sampling period, in the short term (first 1-2 generations), or in the long term (last 1-2 generations), for totals and segregated into different fisheries (e.g., mixed-CU vs. terminal). (Median value over MC trials and 95\\% CIs)",
  "", "Catch variability: CV of (or average \\% change between years in) catch over the sampling period, for totals and segregated into different fisheries (e.g., mixed vs. terminal). (Median value over MC trials and 95\\% CIs)","","",

  "Allocate catch to terminal vs. mixed-CU fisheries", "-Proportion of catch in mixed-CU vs terminal fisheries averaged over the entire sampling period (Median value over MC trials and 95\\% CIs)")

dat %>% 
  mutate_all(kableExtra::linebreak) %>%
  csas_table(format = "latex", escape = FALSE, font_size = 11, linesep = "\\addlinespace",
             caption = "Examples of objectives and performance metric that could be used to rebuild Pacific salmon CUs") %>%
  #kableExtra::row_spec(4, hline_after = TRUE) %>% 
  #kableExtra::row_spec(11, hline_after = TRUE) %>% 
  #kableExtra::row_spec(17, hline_after = TRUE) %>% 
  #kableExtra::row_spec(19, hline_after = TRUE) %>% 
  #kableExtra::row_spec(23, hline_after = TRUE) %>% 
  kableExtra::column_spec(1, width = "15em") %>%
  kableExtra::column_spec(2, width = "30em")
             
             

```



See Table \@ref(tab:testtab) for the example table.

See Figure \@ref(fig:fig1) for the example figure.

\clearpage
